Astrophysical calibration of gravitational-wave detectors

### 2015-04-08 docs/cosmic_cal.tex
Added stuff and tidied things a bit.

\begin{abstract}
  We present an investigation into the potential of calibrating
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmologic
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.
\end{abstract}

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivelent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{NSBH}[NSBH]{neutron star black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}

@@

\cm{Explain the basic idea.}

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see~\cite{Vitale} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the 'hardware injection method' is roughly $10\%$~\cite{Vitale}. This is a
benchmark set for the error estimation using the proposed new method in this
project.

The hardware method of calibrating the ground-based detectors is incredibly
complex in its application, with many sources of error which have to be
accounted for. The new method of calibration proposed and tested in this
project is a lot simpler. The method acts as an external calibrator, to be
carried out using the data collected by the detector, $\tilde{d}(f)$. We assume
the collected data contains some unknown and unaccounted for calibration error.
The uncertainty in $\tilde{d}(f)$ can then be evaluated using methods of
parameterisation already carried out in GW research. This method of
parameterising GW defining variables is then carried out on a known source, a
standard candle as described above, which allows one to measure the calibration
uncertainty in the data set.

One can estimate the error in the calibration using parameter estimation
methods, or more precisely Bayesian inference. In this project the calibration
offset used throughout is simply a scale factor, $s$. This effects the data set
collected by a simple multiplication:

\begin{equation}
  \label{eq:scaledata}
  \tilde{d}_m(f) = \tilde{d}_e(f) \times s
\end{equation}

where $\tilde{d}_m(f)$ is the data-set measured by the detector, and
$\tilde{d}_e(f)$ is the exact data set, with no error. A description of the
collected data is given in section~\ref{sec:cbc}. The scale factor being
parameterised, $s$, represents all known sources of error in the system. This
is not entirely realistic as $s$ is known to be a function of frequency.
Testing a more realistic calibration error is work to be done in the future.

Being able to resolve the calibration scale factor with the least amount of
uncertainty was the aim of this project; to show that this method is
competitive with current techniques. If the calibration error can be resolved
with known uncertainty, GW data analysis can then be carried out with more
certainty. This is useful as an entirely independent method of assessing the
uncertainty in the current calibration method.

@@

A GW detection on its own right would be ground breaking, allowing
astrophysicists to observe the sky above in a whole new spectrum. However, to
increase the scientific return on a direct GW detection, a counterpart EM
observation of the source itself would need to be seen.  There are two
approaches to observing a source both in EM and GW spectrum's~\cite{grb}:

\begin{enumerate}
  \item locating and observing a source with an EM follow-up to
a GW detection
  \item searching for a GW signal generated by a source seen
through EM observations
\end{enumerate}

Both of these methods have their pros and cons. Firstly, an EM follow-up would
be very difficult due to the duration of GW source system detectable by
ground-based interferometers. A neutron star binary system (NS-NS binary) has a
duration of $\sim15$ minutes~\cite{bnstime}.  Errors in the sky position from a
GW detection would also create issues for an EM follow-up. The sky localisation
obtained from the ground-based detectors gives a large field of view for the EM
observatories to search, roughly an area spanning $100$'s of
degrees~\cite{grb}.

The more promising detection approach then is a follow-up GW search from an EM
trigger, especially for the source of GW considered in this paper. The merger
time for NS-NS binaries are known with an uncertainty within a few seconds for
the EM radiation emitted from this system~\cite{grb}.

Identifying the source of GWs redshift would then lead to calculating the
Hubble constant and other cosmological parameters~\cite{Schutz}. The evolution
of the system could be better understood, allowing for searches with greater
accuracy. Counterpart EM observations of a GW source will also remove
uncertainty in the defining GW parameters, a fact which was taken advantage of
for this project.

The source of GW considered in this paper will be a coalescing binary system
consisting of neutron stars. As well as being the most probable source of GWs
detectable by ground-based interferometers, they are also expected to emit EM
radiation. As shown in figure~\ref{fig:grb}, the emission occurs due to the
high velocity with which the stars coalesce. The coalescences are thought to be
the progenerators of short-duration gamma ray bursts (SGRBs)~\cite{grb}.
Observations of this radiation has been see, through the Swift satellite and
follow-up ground-based telescopes~\cite{swift}. Unfortunately, no observation
of SGRBs have been seen within the range of advanced generation of LIGO and
Virgo. It is probable, however, that such a source exists well within the range
of advanced LIGO/ Virgo, but has simply not been observed~\cite{grb}. This then
validates the assumption made throughout the project.

In this paper, a key assumption has to be made in order to test this new method
of calibration. It is assumed that there exists a source of GWs with enough
information known through EM observation such that it can be used as a
\textit{standard candle}, i.e. an accurately known and well modelled source of
GWs. In this project (for the source of GWs being analysed), a GRB observation
is assumed to have been seen. This allows for discrepancies in the detector
calibration itself to be parameterised. This is the proposed new method of
calibration for the detectors, for all types of detectors.

@@

\cm{This should be the simplest case we can do.}

@@

Coalescing binary systems of compact objects, such as neutron stars or black
holes, are the most probable source of GWs detectable by ground-based
interferometers~\cite{peadvanced}. A direct detection of GWs would allow for a
fuller understanding of the mechanics of these systems. Compact binary
coalescence's (CBCs) are systems of two compact object inspiralling, with
reducing separation, until the inevitable merger of the objects. This orbital
motion creates ripples in space-time; this is the GW from this motion. For
gravitational wave research, the interests lie in the GW signal generated, or
the \textit{waveform}, shown in figure~\ref{fig:signal}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{grav_signal}
  \caption{The gravitational waveform produced using coalescing binary
neutron stars. The wave is represented in the frequency domain with masses
$m_{1,2} = 1.4M_{\odot}$ at a luminosity distance $D_{L} = 200 \,Mpc$.}
  \label{fig:signal}
\end{figure}

@@

>>>>

\begin{abstract}
  We present an investigation into the potential of calibrating
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmologic
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.~\cm{Update at end}
\end{abstract}

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivelent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{NSBH}[NSBH]{neutron star black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect such waves emitted from $O(10s)$ of compact binary
coalescences.  One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx
450$ Mpc. Assuming the current best estimates for the cosmological parameters,
this is equivelent to a redshift $z\approx 0.1$. As noted by~\cite{} \ac{CBC}
systems can be used as cosmiological distance markers, otherwise known as
"standard sirens" (analogous to the \ac{EM} standard candles). Standard sirens
allow ...

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} systems, in addition to
emitting detectable \acp{GW}, is also the mechanism for producing \acp{sGRB}.
In this scenario these \ac{EM} events are produced through the ... and result
in tightly beamed emission parallel to the orbital angular momentum vector of
the \ac{BNS} system. Additional evidence for the coincidence of \acp{sGRB} with
\ac{GW} events are the estimated astrophysical rates of both phenomena.
Observations of \acp{sGRB} give a rate of XXX Mpc$^{-3}$yr$^{-1}$ which can be
compared to estimates of \ac{BNS} merger rates of XXX-XXX Mpc$^{-3}$yr$^{-1}$.
This latter range is obtained from population synthesis models together with
knowledge of the X known \ac{BNS} systems in our galaxy.

% talk about the calibration idea
If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it will
be possible to identify the host galaxy of the \ac{BNS} source.  With this
information a spectroscopic redshift can be very accurately obtained. With
knowledge of the redshift, using current best estimates for the Hubble constant
and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy  The \ac{GW} measurement acts as a standard
siren also giving us a direct measurement of the luminosity distance to the source.
The accuracy of such a measurement depends on a number of factors including the
accuracy with which the \ac{GW} detector has been calibrated.  Hence comparison
with the distance estimate from the \ac{sGRB} we can re-calibrate (or validate)
the existing experimentally obtained calibration.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single
coincident \ac{GW}--\ac{sGRB} events and establish the validation power of such
a calibration technique.  In Sec.~\ref{sec:calibration} we briefly summarize
the existing experimental technique for \ac{GW} detector calibration and its
expected accuracy.  We then review the concept of \ac{GW} standard sirens and
their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Sec.~\ref{sec:single} we present
the results of Monte-Carlo simulations for the purpose of luminosity distance
estimation using \ac{GW} events. In Sec.~\ref{sec:cosmo} we show the likely
accuracy of \ac{sGRB} distances obtained from their host galaxy redshift and in
Sec.~\ref{sec:discussion} we conclude.

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see~\cite{Vitale} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the 'hardware injection method' is roughly $10\%$~\cite{Vitale}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

% Describe our model for the calibration
In this work the calibration correction factor used throughout is simply a scale
factor, $s$. This effects the data set collected by a simple multiplication:
%
\begin{equation}\label{eq:scaledata}
  h(t) = s\times h_{\mathrm{m}}(t)
\end{equation}
%
where $h_{\mathrm{m}}(t)$ is the measured detector strain, and
$h(t)$ is the exact perfectly calibrated strain. A description of the
collected data is given in section~\ref{sec:cbc}. The scale factor being
parameterised, $s$, represents all known sources of error in the system. This
is not entirely realistic as $s$ is known to be a function of frequency.
Testing a more realistic calibration error is work to be done in the
future.~\cm{Need to clean this up and expand the discussion}

@@

% what are sGRBs 
\cm{What are sGRBs?}\\

% discuss joint-event rates
\cm{Discuss the expected rate of joint detections}\\

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would
be identified is through the targeted follow-up of an observed \ac{sGRB} or by
post-facto matching of \ac{GW} trigger lists with known \ac{sGRB} events.  The
likelihood of being able to follow-up \ac{GW} events using gamma-ray telescopes
with low enough latency to catch a \ac{sGRB} is low.  Compounding this issue is
the relatively large \ac{GW} sky error-box giving a field-of-view for the
\ac{EM} observatories to search spanning $\sim 100$'s of square
degrees~\cite{grb}.  For the \ac{GW} follow-up of \ac{sGRB} scenario the merger
time for \ac{BNS} systems will be estimated from the \ac{sGRB} to within a few
seconds~\cite{grb}.  This makes the follow-up search less computationally
expensive since it is performed over a smaller range of data using potentially
fewer numbers of waveform templates.  This computational saving enables the use
of a more computationally expensive multi-detector coherent scheme rather than
the cheaper coincidence methods used in the untargeted searches.

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the
fact that in order for such an event to be observed, the \ac{BNS} system must
have had its orbital angular momentum vector pointing towards (or away) from
the detector.  The actual inclination angle of the system, defined as the angle
between the orbital angular momentum vector and the line of sight, must be $<$
half of the \ac{sGRB} beaming angle.  This prerequisite property limits us to
systems that are approximately "face-on" and therefore biases us to higher
\ac{SNR} signals.  However, as discussed earlier, the property of beaming
severely impacts the probable rate of such joint observations.

@@

\cm{This should be the simplest case we can do.  We need to cut a lot of this.}

% Describe what we are going to do in this section

% Describe the signal model (Taylor F2 etc...)

% Define the detectors and the noise epoch we will be using

% Define the likelihood function we will be using

% describe what parameters we will be considering

% describe the priors

% mention the no-noise issue

% Describe the ensemble of simulations used

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Distance estimates from \acp{sGRB}\label{sec:cosmo}}

~\cm{Describe the cosmological analysis used to obtain the distance from the
GRB}
